RDF Validation Requirements - Evaluation and Logical Underpinning

### 2015-01-06 paper/RDF validation requirements/RDF validation requirements.tex
add descriptions

\begin{itemize}
	\item R-13-DISJOINT-GROUP-OF-PROPERTIES-CLASS-SPECIFIC
\end{itemize}

\begin{itemize}
	\item exclusive or: logical operation that outputs true whenever both inputs differ (one is true, the other is false).
\end{itemize}

@@

\begin{ex}
# DSCL (ShEx)
# -----------
<Human> { 
    (  
        foaf:name xsd:string ,
        foaf:givenName xsd:string ) 
    |
    (
        foaf:mbox IRI ,
        foaf:homepage foaf:Document ) }
\end{ex}

@@

\begin{ex}
# valid data
# ----------
:Thomas
    a :Human ;
    foaf:mbox <thomas.bosch@gesis.org> ;
    foaf:homepage <http://purl.org/net/thomasbosch> .
\end{ex}

\begin{ex}
# invalid data
# ----------
:Thomas
    a :Human ;
    foaf:name 'Thomas Bosch' ;
    foaf:givenName 'Thomas' ;
    foaf:mbox <thomas.bosch@gesis.org> ;
    foaf:homepage <http://purl.org/net/thomasbosch> .
\end{ex}

@@

\begin{ex}
# ShExC
# -----
[[
  <UserShape> {                          # A User has
    (foaf:name xsd:string                #  either a foaf:name
     | foaf:givenName xsd:string+,       #   or 1+ givenNames
       foaf:familyName xsd:string),      #   and a family name
    foaf:mbox IRI                        # and an mbox.
  }
]]
\end{ex}

\begin{ex}
# ReSh
# ----

[[
  PREFIX rs:<http://open-services.net/ns/core#>
  PREFIX se:<http://www.w3.org/2013/ShEx/Definition#>

  <UserShape> a rs:ResourceShape ;       # A User has
    se:choice [
      rs:property [                      #  either a foaf:name
        rs:name "name" ;
        rs:propertyDefinition foaf:name ;
        rs:valueType xsd:string ;
        rs:occurs rs:Exactly-one ;
      ] ;
      se:ruleGroup [
        rs:property [                    #   or 1+ givenNames
          rs:name "givenName" ;
          rs:propertyDefinition foaf:givenName ;
          rs:valueType xsd:string ;
          rs:occurs rs:One-or-many ;
        ] ;
        rs:property [                    #   and a family name
          rs:name "familyName" ;
          rs:propertyDefinition foaf:familyName ;
          rs:valueType xsd:string ;
          rs:occurs rs:Exactly-one ;
        ] ;
      ] ;
    ] ;
    rs:property [                        # and an mbox.
      rs:name "mbox" ;
      rs:propertyDefinition foaf:mbox ;
      rs:valueType rs:Resource ;
      rs:occurs rs:Exactly-one ;
    ] .
]]
\end{ex}

@@

\begin{itemize}
  \item at least one property group must match for individuals of a specific context 
	\item context: may be a class
\end{itemize}

@@

\begin{ex}
# valid data
# ----------
:Anakin
    a :Person ;
    foaf:firstName 'Anakin' ;
    foaf:lastName 'Skywalker' ;
    foaf:givenName 'Anakin' ;
    foaf:familyName 'Skywalker' .
\end{ex}

\begin{ex}
# invalid data
# ----------
:Anakin
    a :Person .
\end{ex}

@@

syntactic sugar:
\begin{gcotable}
class & Human & - & - & E, F & $\sqcup$ \\
class & E & - & - & = 1 name.string, = 1 givenName.string & $\sqcap$ \\
class & F & - & - & = 1 mbox.IRI, = 1 homepage.Document & $\sqcap$ \\
property & = 1 name.string & name & - & string & = & 1 \\
property & = 1 givenName.string & givenName & - & string & = & 1 \\
property & = 1 mbox.IRI & mbox & - & IRI & = & 1 \\
property & = 1 homepage.Document & homepage & - & Document & = & 1 \\
\end{gcotable}

@@

\begin{itemize}
	\item R-30-ALLOWED-VALUES-FOR-RDF-OBJECTS
  \item R-37-ALLOWED-VALUES-FOR-RDF-LITERALS
\end{itemize}

\begin{itemize}
  \item It is a common requirement to narrow down the value space of a property by an exhaustive enumeration of the valid values (both literals or resource). This is often rendered in drop down boxes or radio buttons in user interfaces. 
	\item define a concept by simply enumerating its instances
	\item A DL nominal is a concept that has exactly one instance
	\item DL enumeration: combining nominals with union
\end{itemize}

allowed values for properties

\begin{itemize}
	\item must be an IRI
  \item must be an IRI matching this pattern (e.g. http://id.loc.gov/authorities/names/)
  \item must be an IRI matching one of these patterns
  \item must be a (any) literal
  \item must be one of these literals ("red" "blue" "green")
  \item must be a typed literal of this type (e.g. XML dataType)
\end{itemize}

@@

\begin{ex}
# DSP
# ----------
:descriptionTemplate 
    a dsp:DescriptionTemplate ;
    dsp:minOccur "0"^^xsd:nonNegativeInteger ; 
    dsp:maxOccur "infinity"^^xsd:string ; 
    dsp:resourceClass swrc:Book ; 
    dsp:statementTemplate [
        a dsp:NonLiteralStatementTemplate ;
        dsp:minOccur "0"^^xsd:nonNegativeInteger ; 
        dsp:maxOccur "infinity"^^xsd:string ; 
        dsp:property dcterms:subject ; 
        dsp:nonLiteralConstraint [ 
            a dsp:NonLiteralConstraint ;
            dsp:valueClass skos:Concept ;
            dsp:valueURI :ComputerScience, :SocialScience, :Librarianship ] ] .
\end{ex}

\begin{ex}
# OWL2
# ----
dcterms:subject rdfs:range :ObjectOneOf . 
# EquivalentClasses( :ObjectOneOf ObjectOneOf( :ComputerScience :SocialScience :Librarianship ) )
:ObjectOneOf owl:equivalentClass [ 
    a owl:Class ;
    owl:oneOf ( :ComputerScience :SocialScience :Librarianship ) ] .
\end{ex}

\begin{itemize}
	\item the range of the object property dcterms:subject must consist of the individuals :ComputerScience :SocialScience :Librarianship which are of the class skos:Concept
\end{itemize}

\begin{DL}
$\top$ $\sqsubseteq$ $\forall$ dcterms:subject . skos:Concept $\sqcap$ ( $\{ComputerScience\}$ $\sqcup$ \{SocialScience\} $\sqcup$ \{Librarianship\} ) \\
\end{DL}

@@

\begin{itemize}
	\item Jedis have blue, green, or white laser swords
\end{itemize}

\begin{ex}
# DSP
# ---
:personDescriptionTemplate
    a dsp:DescriptionTemplate ;
    dsp:resourceClass :Jedi ;
    dsp:statementTemplate [
        a dsp:LiteralStatementTemplate ;
        dsp:property :laserSwordColor ;
        dsp:literalConstraint [
            a dsp:LiteralConstraint ;
            dsp:literal "blue" ;
            dsp:literal "green" ;
            dsp:literal "white"] ] .
\end{ex}

\begin{ex}
# OWL2
# ----
:laserSwordColor rdfs:range :laserSwordColors . 
:laserSwordColors
    a rdfs:Datatype .
    owl:oneOf ( "blue" "green" "white" ) .
\end{ex}

\begin{ex}
# ReSh
# ----
:Jedi a rs:ResourceShape ;
    rs:property [
        rs:name "laserSwordColor" ;
        rs:propertyDefinition :laserSwordColor ;
        rs:allowedValue "blue" , "green" , "white" ;
        rs:occurs rs:Exactly-one ;
    ] .
\end{ex}

\begin{ex}
# ShEx
# ----
:Jedi {
    :laserSwordColor ('blue' 'green' 'white') }
\end{ex}

\begin{ex}
# Jedi individuals
# ----------------
:Yoda 
    :laserSwordColor 'blue' .
\end{ex}

\begin{DL}
Jedi $\equiv$ $\exists$ laserSwordColor . \{blue, green, white\} \\
\end{DL}

>>>>

\ms{Context-Specific Exclusive OR of Property Groups} corresponds to the requirement:
\ms{R-13-DISJOINT-GROUP-OF-PROPERTIES-CLASS-SPECIFIC}.
Exclusive or is a logical operation that outputs true whenever both inputs differ (one is true, the other is false).
Only one of multiple property groups leads to valid data.

@@

\begin{ex}
# ShEx:
<Human> { 
    (  
        foaf:name xsd:string ,
        foaf:givenName xsd:string ) 
    |
    (
        foaf:mbox IRI ,
        foaf:homepage foaf:Document ) }
\end{ex}

@@

\begin{ex}
# valid data:
Thomas
    a Human ;
    foaf:mbox <thomas.bosch@gesis.org> ;
    foaf:homepage <http://purl.org/net/thomasbosch> .
\end{ex}

\begin{ex}
# invalid data:
Thomas
    a Human ;
    foaf:name 'Thomas Bosch' ;
    foaf:givenName 'Thomas' ;
    foaf:mbox <thomas.bosch@gesis.org> ;
    foaf:homepage <http://purl.org/net/thomasbosch> .
\end{ex}

@@

%\subsection{Complex Example}
%
%\begin{ex}
%# ShExC
%# -----
%[[
  %<UserShape> {                          # A User has
    %(foaf:name xsd:string                #  either a foaf:name
     %| foaf:givenName xsd:string+,       #   or 1+ givenNames
       %foaf:familyName xsd:string),      #   and a family name
    %foaf:mbox IRI                        # and an mbox.
  %}
%]]
%\end{ex}
%
%\begin{ex}
%# ReSh
%# ----
%
%[[
  %PREFIX rs:<http://open-services.net/ns/core#>
  %PREFIX se:<http://www.w3.org/2013/ShEx/Definition#>
%
  %<UserShape> a rs:ResourceShape ;       # A User has
    %se:choice [
      %rs:property [                      #  either a foaf:name
        %rs:name "name" ;
        %rs:propertyDefinition foaf:name ;
        %rs:valueType xsd:string ;
        %rs:occurs rs:Exactly-one ;
      %] ;
      %se:ruleGroup [
        %rs:property [                    #   or 1+ givenNames
          %rs:name "givenName" ;
          %rs:propertyDefinition foaf:givenName ;
          %rs:valueType xsd:string ;
          %rs:occurs rs:One-or-many ;
        %] ;
        %rs:property [                    #   and a family name
          %rs:name "familyName" ;
          %rs:propertyDefinition foaf:familyName ;
          %rs:valueType xsd:string ;
          %rs:occurs rs:Exactly-one ;
        %] ;
      %] ;
    %] ;
    %rs:property [                        # and an mbox.
      %rs:name "mbox" ;
      %rs:propertyDefinition foaf:mbox ;
      %rs:valueType rs:Resource ;
      %rs:occurs rs:Exactly-one ;
    %] .
%]]
%\end{ex}

@@

At least one property group must match for individuals of a specific context. 
Context may be a class, a shape, or an application profile.

@@

\begin{ex}
# valid data:
:Anakin
    a :Person ;
    foaf:firstName 'Anakin' ;
    foaf:lastName 'Skywalker' ;
    foaf:givenName 'Anakin' ;
    foaf:familyName 'Skywalker' .
\end{ex}

\begin{ex}
# invalid data:
:Anakin
    a :Person .
\end{ex}

@@

\begin{gcotable}
class & Human & - & - & E, F & $\sqcup$ \\
class & E & - & - & = 1 name.string, = 1 givenName.string & $\sqcap$ \\
class & F & - & - & = 1 mbox.IRI, = 1 homepage.Document & $\sqcap$ \\
property & = 1 name.string & name & - & string & = & 1 \\
property & = 1 givenName.string & givenName & - & string & = & 1 \\
property & = 1 mbox.IRI & mbox & - & IRI & = & 1 \\
property & = 1 homepage.Document & homepage & - & Document & = & 1 \\
\end{gcotable}

@@

\ms{Allowed Values} corresponds to the requirements:
\ms{R-30-ALLOWED-VALUES-FOR-} \ms{RDF-OBJECTS} and 
\ms{R-37-ALLOWED-VALUES-FOR-RDF-LITERALS}.
It is a common requirement to narrow down the value space of a property by an exhaustive enumeration of the valid values (both literals or resource). This is often rendered in drop down boxes or radio buttons in user interfaces. 
%\begin{itemize}
  %\item define a concept by simply enumerating its instances
	%\item A DL nominal is a concept that has exactly one instance
	%\item DL enumeration: combining nominals with union
%\end{itemize}
Allowed values for properties
\begin{itemize}
	\item must be these IRIs,
  \item must be IRIs matching specific patterns,
  \item must be IRIs matching one of multiple patterns,
  \item must be (any) literals,
  \item must be literals of a list of allowed literals (e.g. "red" "blue" "green"),
  \item must be typed literals of this type (e.g. XML dataType).
\end{itemize}

@@

\begin{ex}
# DSP:
descriptionTemplate 
    a dsp:DescriptionTemplate ;
    dsp:minOccur "0"^^xsd:nonNegativeInteger ; 
    dsp:maxOccur "infinity"^^xsd:string ; 
    dsp:resourceClass swrc:Book ; 
    dsp:statementTemplate [
        a dsp:NonLiteralStatementTemplate ;
        dsp:minOccur "0"^^xsd:nonNegativeInteger ; 
        dsp:maxOccur "infinity"^^xsd:string ; 
        dsp:property dcterms:subject ; 
        dsp:nonLiteralConstraint [ 
            a dsp:NonLiteralConstraint ;
            dsp:valueClass skos:Concept ;
            dsp:valueURI ComputerScience, SocialScience, Librarianship ] ] .
\end{ex}

\begin{ex}
# OWL2:
dcterms:subject rdfs:range ObjectOneOf . 
# EquivalentClasses( ObjectOneOf ObjectOneOf( ComputerScience SocialScience Librarianship ) )
ObjectOneOf owl:equivalentClass [ 
    a owl:Class ;
    owl:oneOf ( ComputerScience SocialScience Librarianship ) ] .
\end{ex}

The range of the object property \ms{dcterms:subject} must consist of the individuals \ms{ComputerScience} \ms{SocialScience} \ms{Librarianship} which are of the class \ms{skos:Concept}:

\begin{DL}
$\top$ $\sqsubseteq$ $\forall$ dcterms:subject . skos:Concept $\sqcap$ \\
( $\{ComputerScience\}$ $\sqcup$ \{SocialScience\} $\sqcup$ \{Librarianship\} ) \\
\end{DL}

@@

Jedis have blue, green, or white laser swords.

\begin{ex}
# DSP:
personDescriptionTemplate
    a dsp:DescriptionTemplate ;
    dsp:resourceClass Jedi ;
    dsp:statementTemplate [
        a dsp:LiteralStatementTemplate ;
        dsp:property laserSwordColor ;
        dsp:literalConstraint [
            a dsp:LiteralConstraint ;
            dsp:literal "blue" ;
            dsp:literal "green" ;
            dsp:literal "white"] ] .
\end{ex}

\begin{ex}
# OWL2:
laserSwordColor rdfs:range laserSwordColors . 
laserSwordColors
    a rdfs:Datatype .
    owl:oneOf ( "blue" "green" "white" ) .
\end{ex}

\begin{ex}
# ReSh:
Jedi a rs:ResourceShape ;
    rs:property [
        rs:name "laserSwordColor" ;
        rs:propertyDefinition laserSwordColor ;
        rs:allowedValue "blue" , "green" , "white" ;
        rs:occurs rs:Exactly-one ; ] .
\end{ex}

\begin{ex}
# ShEx:
Jedi {
    laserSwordColor ('blue' 'green' 'white') }
\end{ex}

\begin{ex}
# Jedi individuals:
Yoda 
    laserSwordColor 'blue' .
\end{ex}

\begin{DL}
Jedi $\equiv$ $\exists$ laserSwordColor.\{blue, green, white\} \\
\end{DL}
